Fusing Filters with Integer Linear Programming

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
Sec 4 (sole transducer as def. for val. constr)

Although our constraint based formulation of size inference is reminiscent of type inference for HM(X)~\cite{odersky1999type}, there are important differences. Firstly, our type schemes include existential quantifiers, which represent the fact that the sizes of arrays produced by filter operations are unknown in general. This is also the case for @generate@, where the result size is data dependent. HM(X) style type inferences uses the $\exists$ quantifier to bind local type variables in constraints, and existential quantifiers do not appear in type schemes. Secondly, our types are first order only, as program graphs cannot take other program graphs as arguments. Provided we generate the constraints in the correct form, solving them is straightforward.

@@

Note that the constraints are generated in a specific form, to make the constraint solving process easy. For each array variable in the program, we generate a new size variable, like size $k_{zs}$ for array variable $zs$. These new size variables always appear on the \emph{left} of atomic equality constraints. For each array binding we may also introduce unification or rigid variables, and these appear on the \emph{right} of atomic equality constraints.

@@

% -------------------------------------------------------------------
\subsection{Constraint Solving and Generalization}
The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a function by first extracting size constraints, before solving them and generalizing the result. In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes and environment and constraint, producing a solved environment and constraint. As the constraint solving process is both standard and straightforward, we only describe it informally.

Recall from the previous section that in our generated constraints all the size variables named after program variables are on the left of atomic equality constraints, while all the unification and existential variables are on the right. To solve the constraints we keep finding pairs of atomic equality constraints where the same variable appears on the left, unify the right of both of these constraints, and apply the resulting substitution to both the environment and original constraints. When there are no more pairs of constraints with the same variable on the left then the constraints are in solved form and we are finished.

During constraint solving, all unification variables mentioned in the environment can have other sizes substituted for them. In contrast, the rigid variables marked by the $\exists$ symbol cannot. For example, if we consider the constraints for @normalize2@ mentioned before:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & \true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  \true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately. This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}. In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Note that, although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule. The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another. To say this another way, all subgraphs are fully inlined. Recall from \S\ref{s:CombinatorNormalForm} that we assume our operator graphs are embedded in a larger host program. We use size information to guide the clustering process, and although the host program can certainly call the operator graph, static size information does not flow across this boundary.

@@

We use $T$ to range over iteration sizes, and write $\bot$ for the case where the iteration size is unknown. The $\bot$ size is needed to handle the @external@ operator, as in general we cannot know its true iteration size, and it cannot be fused with any other operator.

@@

Once the size constraints have been solved, we can use the following function to compute the iteration size of each binding. In the definition we use the syntax $\Gamma(xs)$ to find the $xs : k$ element in the environment $\Gamma$ and return the associated size $k$. Similarly, we use the syntax $C(k)$ to find the corresponding $k = \tau$ constraint in $C$ and return the associated size type $\tau$.

@@

% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
% No,
We now define a function $trans$, to find the parent transducer a combinator application. Since each name is bound to at most one combinator, we abuse terminology here slightly and write ``combinator $n$'' when refering to the combinator occuring in the binding of the name $n$.
The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.

@@

Two combinators $a$ and $b$ of different size may be fused together only if they have parents $(c, d) \in parents(a,b)$, and the combinators and their parents are also fused together.
That is, in order for $a$ and $b$ to be fused together, $c$ and $d$ must be fused, $a$ and $c$ must be fused, and $d$ and $b$ must be fused.
In the previous example, @sum1@ and @sum2@ have different iteration size, and their parents are $parents(@sum1@, @sum2@) = \{(@sum1@, @gts@)\})$.
In order for @sum1@ and @sum2@ to be fused together, @sum1@ and @gts@ must be fused, @sum1@ and @sum1@ must be fused, and @gts@ and @sum2@ must be fused.

\textbf{Lemma: sole transducers}.
For some function $f$ with valid constraints, each name $n$ will have at most one transducer.
\[
\forall f, \sigma, n.\ f :_s \sigma \implies |trans(binds(f), n)| \le 1
\]
% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
% The other cases are trivial.

@@

These two lemmas are used in the integer linear programming formulation, when constructing the program.
When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

>>>>

Although our constraint based formulation of size inference is reminiscent of type inference for HM(X)~\cite{odersky1999type}, there are important differences. Firstly, our type schemes include existential quantifiers, which express the fact that the sizes of arrays produced by filter operations are unknown in general. This is also the case for @generate@, where the result size is data dependent. HM(X) style type inferences use the $\exists$ quantifier to bind local type variables in constraints, and existential quantifiers do not appear in type schemes. Secondly, our types are first order only, as program graphs cannot take other program graphs as arguments. Provided we generate the constraints in the correct form, solving them is straightforward.

@@

Note that the constraints are generated in a specific form, to facilitate the constraint solving process. For each array variable in the program, we generate a new size variable, like size $k_{zs}$ for array variable $zs$. These new size variables always appear on the \emph{left} of atomic equality constraints. For each array binding we may also introduce unification or rigid variables, and these appear on the \emph{right} of atomic equality constraints.

@@

% -------------------------------------------------------------------
\subsection{Constraint Solving and Generalization}
The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a function by first extracting size constraints, before solving them and generalizing the result. In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes an environment and a constraint set, and produces a solved environment and constraint set. As the constraint solving process is both standard and straightforward, we only describe it informally.

Recall from the previous section that in our generated constraints all the size variables named after program variables are on the left of atomic equality constraints, while all the unification and existential variables are on the right. To solve the constraints, we keep finding pairs of atomic equality constraints where the same variable appears on the left, unify the right of both of these constraints, and apply the resulting substitution to both the environment and original constraints. When there are no more pairs of constraints with the same variable on the left, the constraints are in solved form and we are finished.

During constraint solving, all unification variables occuring in the environment can have other sizes substituted for them. In contrast, the rigid variables marked by the $\exists$ symbol cannot. For example, consider the constraints for @normalize2@ mentioned before:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & \true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  \true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

To produce the final size scheme, we lookup the sizes of the input and output variables of the original function from the solved constraints and generalize appropriately. This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}. In this case, no rigid size variables appear in the result, so we can universally quantify all size variables.
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Note that, although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule. The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another. To say this another way, all subgraphs are fully inlined. Recall from \S\ref{s:CombinatorNormalForm}, that we assume our operator graphs are embedded in a larger host program. We use size information to guide the clustering process, and although the host program can certainly call the operator graph, static size information does not flow across this boundary.

@@

We use $T$ to range over iteration sizes, and write $\bot$ for the case where the iteration size is unknown. The $\bot$ size is needed to handle the @external@ operator, as we cannot statically infer its true iteration size, and it cannot be fused with any other operator.

@@

Once the size constraints have been solved, we can use the following function to compute the iteration size of each binding. In the definition, we use the syntax $\Gamma(xs)$ to find the $xs : k$ element in the environment $\Gamma$ and return the associated size $k$. Similarly, we use the syntax $C(k)$ to find the corresponding $k = \tau$ constraint in $C$ and return the associated size type $\tau$.

@@

% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
% No,
We now define a function $trans$, to find the parent transducer a combinator application. Since each name is bound to at most one combinator, we abuse terminology here slightly and write \emph{combinator $n$} when refering to the combinator occuring in the binding of the name $n$.
The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.

@@

Two combinators $a$ and $b$ of different size may be fused together only if they have parents $(c, d) \in parents(a,b)$, and the combinators and their parents are also fused together.
That is, in order for $a$ and $b$ to be fused together, $c$ and $d$ must be fused, $a$ and $c$ must be fused, and $d$ and $b$ must be fused.
In the previous example, @sum1@ and @sum2@ have different iteration size, and their parents are $parents(@sum1@, @sum2@) = \{(@sum1@, @gts@)\})$.
In order for @sum1@ and @sum2@ to be fused together, @sum1@ and @gts@ must be fused, @sum1@ and @sum1@ must be fused, and @gts@ and @sum2@ must be fused. Now we can express the restriction on programs we view as valid for our transformation more formally:

%\textbf{Lemma: sole transducers}.
\TODO{Double check: is this really a sufficient property, or only necessary?}
\textbf{Definition: valid constraints}.
The constraints of a function $f$ are \emph{valid}, if and only if each name $n$ in its bindings will have at most one transducer:
\[
\forall f, \sigma, n.\ f :_s \sigma \implies |trans(binds(f), n)| \le 1
\]
% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
% The other cases are trivial.

@@

These two properties are used in the integer linear programming formulation, when constructing the program.
When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.
